CFTP Convergence

### 2025-01-10 application_examples/abc_inferences/archipelago-definition.Rmd
Updating git in order to pull

```{r}
# Extract TSS
promoters <- data.frame(
  chr = seqnames(gtf_transcripts),
  tss = ifelse(strand(gtf_transcripts) == "+", start(gtf_transcripts), end(gtf_transcripts)),
  strand = strand(gtf_transcripts)
)
head(promoters)
```

The promoters data frame contains:

@@

```{r}

@@

>>>>

```{r}
# Extract TSS
tss <- data.frame(
  chr = seqnames(gtf_transcripts),
  tss = ifelse(strand(gtf_transcripts) == "+", start(gtf_transcripts), end(gtf_transcripts)),
  strand = strand(gtf_transcripts)
)
head(tss)
```

The tss data frame contains:

@@

# Dirk idea

## Explore the data

```{r}
unique(gtf$type)
```

The gtf contains: gene, transcript, exon, CDS, start_codon and stop_codon

Lets filter the genes:

```{r}
genes <- gtf[gtf$type == "gene",]
dim(as.data.frame(genes))
```

Are the genes duplicated?

```{r}
length(genes$gene_id)
length(unique(genes$gene_id))
```

Lets explore the annotation sources:

```{r}
unique(genes$source)
```

Summary table:

| Source        | Primary Focus                       | Reliability          | Features                                         |
|---------------|-------------------------------------|----------------------|--------------------------------------------------|
| Gnomon        | Computational gene prediction       | Medium               | Predicted genes, transcripts, and exons         |
| cmsearch      | RNA homology                        | Medium-High          | Non-coding RNAs (e.g., ribosomal RNAs)          |
| tRNAscan-SE   | tRNA genes                          | High                 | tRNA annotations                                 |
| BestRefSeq    | High-confidence curated annotations | High                 | Protein-coding and non-coding genes             |
| BestRefSeq, Gnomon | Combined curated and predicted annotations | Medium-High | Hybrid of RefSeq and Gnomon predictions         |
| RefSeq        | Broader RefSeq annotations          | Mixed (High to Low)  | Protein-coding and non-coding genes             |

## Define the 200 kb upstream regions

First, save chromosome lengths:

```{r}
gbff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gbff" 
file_lines <- readLines(gbff_file)
# Extract the LOCUS line
locus_lines <- file_lines[grep("^LOCUS", file_lines)]
head(locus_lines)
```

They dont seem to end with the dot and the number (version) as in the case of the `gtf_genes` data. Lets check if the `gtf_genes` (or the dataframe obtained from it) has several versions:

```{r}
length(unique(genes_df$seqnames))
# Remove version suffix from seqnames
length(unique(sub("\\.\\d+$", "", genes_df$seqnames)))
```

It does not have several versions.

```{r}
# Extract chromosome names and lengths
chrom_info <- gsub("^LOCUS\\s+(\\S+)\\s+(\\d+)\\s+.*$", "\\1,\\2", locus_lines)
chrom_info <- strsplit(chrom_info, ",")
chrom_names <- sapply(chrom_info, `[`, 1)
chrom_lengths <- as.numeric(sapply(chrom_info, `[`, 2))
# Create a named vector
chrom_lengths <- setNames(chrom_lengths, chrom_names)

# Remove version numbers from seqnames in genes_df
genes_df$seqnames <- sub("\\.\\d+$", "", genes_df$seqnames)

@@

Calculate the upstream start and end to prevent overlap with the previous gene.

```{r}
# Ensure genes_df is sorted by seqnames and start position
genes_df <- genes_df[order(genes_df$seqnames, genes_df$start), ]

# Initialize the upstream_start column
genes_df$upstream_start <- NA

# Iterate over genes to adjust upstream_start
for (i in 1:nrow(genes_df)) {
  current_gene <- genes_df[i, ]
  current_chrom <- current_gene$seqnames
  current_strand <- current_gene$strand
  current_start <- current_gene$start
  current_end <- current_gene$end

  # Find the previous gene on the same chromosome
  previous_gene <- genes_df[i - 1, ]
  if (i > 1 && previous_gene$seqnames == current_chrom) {
    previous_end <- previous_gene$end
    previous_start <- previous_gene$start
  } else {
    previous_end <- NA
    previous_start <- NA
  }

  # Adjust upstream_start based on strand and previous gene position
  if (current_strand == "+") {
    if (!is.na(previous_end)) {
      genes_df$upstream_start[i] <- max(current_start - 200000, previous_end)
    } else {
      genes_df$upstream_start[i] <- max(current_start - 200000, 1)
    }
  } else if (current_strand == "-") {
    if (!is.na(previous_start)) {
      genes_df$upstream_start[i] <- min(current_end + 200000, previous_start)
    } else {
      genes_df$upstream_start[i] <- min(current_end + 200000, chrom_lengths[current_chrom])
    }
  }
}

Filter for "curated" chromosomes:

```{r}
# Filter genes_df to keep only those rows where seqnames starts with "NC"
genes_df_filtered <- genes_df[grep("^NC", genes_df$seqnames), ]

# Check the result
head(genes_df_filtered)

Still it seems there is nothing but genes. Maybe I should have filtered for something else.

In a GTF (Gene Transfer Format) file, the gene feature type represents the entire gene, including all its exons, introns, and regulatory regions. The transcript feature type corresponds to a specific transcript variant of the gene, encompassing all exons and introns of that transcript. The exon feature type denotes individual exonic regions within a transcript. The CDS (Coding Sequence) feature type indicates the portion of the gene or transcript that codes for protein. The start_codon and stop_codon feature types mark the beginning and end of the coding sequence, respectively. 
**Try gff**

### 2025-01-10 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
Updating git in order to pull

```{r, fig.align='center',  out.width='50%'}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_01.pdf")
```

```{r, fig.align='center',  out.width='50%'}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_02.pdf")
```

```{r, fig.align='center',  out.width='50%'}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_03.pdf")
```

```{r, fig.align='center',  out.width='50%'}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_04.pdf")
```

```{r, fig.align='center',  out.width='50%'}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_05.pdf")
```

```{r, fig.align='center',  out.width='50%'}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_06.pdf")
```

```{r, fig.align='center',  out.width='50%'}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_07.pdf")
```

```{r, fig.align='center',  out.width='50%'}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_08.pdf")
```

```{r, fig.align='center',  out.width='50%'}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_09.pdf")
```

```{r, fig.align='center',  out.width='50%'}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_10.pdf")
```

>>>>

```{r, out.width="0.8\\linewidth", include=TRUE, fig.align="center", fig.cap=c("your caption"), echo=FALSE}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_01.pdf")
```
